---
title: "Machine Learning Course Project"
author: "aponici"
date: "February 22, 2016"
output: html_document
---

# Project Assigment

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Load Libraries & Import data

In the following code, I loaded the required libraries, and downloaded the csv data off of the web.

```{r}
# Load libraries
library(abind)
library(arm)
library(caret)
library(kernlab)
library(randomForest)
library(rpart)
library(e1071)

set.seed(1)

# Import data
urlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlTest <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

csvTrain <- "training.csv"
csvTest <-  "testing.csv"

## Treat NA, #DIV/0 and blanks as missing values
if (file.exists(csvTrain)) {
  train <- read.csv(csvTrain, na.strings=c("NA","#DIV/0!",""))
} else { 
  download.file(urlTrain,csvTrain)
  train <- read.csv(csvTrain, na.strings=c("NA","#DIV/0!",""))
}                           

if (file.exists(csvTest)) {
  test <- read.csv(csvTest, na.strings=c("NA","#DIV/0!",""))
} else { 
  download.file(urlTest,csvTest)
  test <- read.csv(csvTest, na.strings=c("NA","#DIV/0!",""))
}
```

# Evaluate data

Taking a look at the data, it is immediately apparent that there is a great deal of variables and observations. The first primary concern would be making sure that the data is preprocessed appropriately to have a training set that leads to good predictions.

```{r, eval = FALSE}
# Examine data
summary(test)
summary(train)
```

# Preprocessing data

## Remove extraneous variables

nearZeroVar() is a function that is useful for identifying variables that have zero to little variance. Those variables are removed.

```{r}
# Zero variance variables identified
nearzero <- nearZeroVar(train, saveMetrics = TRUE)
train <- train[, !nearzero$nzv]
```

Next, variables that contain more than 50% missing values are removed as well.

```{r}
toBeRemoved <- sapply(colnames(train), function(x) if(sum(is.na(train[, x])) > 0.50*nrow(train)) 
  {return(TRUE)} else {return(FALSE)}
)
train <- train[, !toBeRemoved]
train <- train[, -(1:6)]
```

## Pre-processing of predictors

```{r}
Hcorr <- caret::findCorrelation(cor(train[, -53]), cutoff=0.8)
names(train)[Hcorr]
```

There are 12 variables that are highly correlative. Principal Component Analysis (PCA) seems to be the best option, because it will help reduce the large set of data to extract a reduced set of principal components.

## Model Specification

In order to avoid overfitting and to reduce out of sample errors, TrainControl is used to perform 7-fold cross validation.

```{r}
tc <- trainControl(method = "cv", number = 7, verboseIter=FALSE , preProcOptions="pca", allowParallel=TRUE)
```

Six models are estimated: Random forest, Support Vector Machine (both radial and linear), a Neural net, a Bayes Generalized linear model and a Logit Boosted model.

```{r}
rf <- train(classe ~ ., data = train, method = "rf", trControl= tc)
svmr <- train(classe ~ ., data = train, method = "svmRadial", trControl= tc)
NN <- train(classe ~ ., data = train, method = "nnet", trControl= tc, verbose=FALSE)
svml <- train(classe ~ ., data = train, method = "svmLinear", trControl= tc)
bayesglm <- train(classe ~ ., data = train, method = "bayesglm", trControl= tc)
logitboost <- train(classe ~ ., data = train, method = "LogitBoost", trControl= tc)
```

Accuracy Comparision

```{r}
model <- c("Random Forest", "SVM (radial)","LogitBoost","SVM (linear)","Neural Net", "Bayes GLM")
Accuracy <- c(max(rf$results$Accuracy),
        max(svmr$results$Accuracy),
        max(logitboost$results$Accuracy),
        max(svml$results$Accuracy),
        max(NN$results$Accuracy),
        max(bayesglm$results$Accuracy))
        
Kappa <- c(max(rf$results$Kappa),
        max(svmr$results$Kappa),
        max(logitboost$results$Kappa),
        max(svml$results$Kappa),
        max(NN$results$Kappa),
        max(bayesglm$results$Kappa))  

performance <- cbind(model,Accuracy,Kappa)

knitr::kable(performance)
```
Random forest and SVM(radial) provide the best results and will provide the predictions for the submission. Even if the Out of sample error cannot be estimated exactly, the in-sample error obtained through cross-validation is calculated over different test sets and should provide a better estimate of out-of sample error with respect to the case of no cross-validation.

Prediction of “classe” variable for the test set
```{r}
rfPred <- predict(rf, test)
svmrPred <- predict(svmr, test)
```

Checking if the models give same predictions
```{r}
prediction <- data.frame(cbind(rfPred, svmrPred))
prediction$same <- with(prediction, rfPred == svmrPred)
colnames(prediction) <- c("Random Forest", "SVM (radial)", "Same Prediction")

knitr::kable(prediction)
```

Generation of the files to be submitted is made through the provided function
```{r}
pml_write_files = function(x){
        n = length(x)
        for(i in 1:n){
                filename = paste0("problem_id_",i,".txt")
                write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
        }
}

pml_write_files(rfPred)
pml_write_files(svmrPred)
```

# Conclusion
The random forest model provides an outstanding accuracy and, accordingly, the predictions for the test set were correct in 100% of the cases.